Molecular dynamics simulation of synchronization in driven particles

### 2021-01-22 manuscript/manuscript_ajp_2020.tex
started on the differential equations

  \item {\bf Integration Methods}
    \label{ex:euler}
    To calculate the position of the particle we
    integrate the equation of motion using
    the standard definition of velocity
    $\vec{v}_i = d\vec{r}_i/dt$ 
    via the 
    %In our model we use the
    Euler method. % to solve this equation.
    Our equation of motion provides
    a direct calculation for particle velocity.
    %
  When solving ordinary differential equations,
  the Euler method is effective for solving linear equations
  of the form
  $dy/dt = f(t,y(t))$ with initial condition $y(t_0) = y_0$.
  The solution is calculated algorithmically
  by stepping in time through $n$ integer steps
  $t_n = t_0 + n \Delta t$.
  At each subsequent step the new
  value for $y$ is calculated as a map solution using
  discrete times 
  $y_{n+1} = y_n + f(t_n,y_n)$.
  The Euler method can be applied to non-linear
  equations effectively if the time step $\Delta t$
  is kept sufficiently small \cite{Newman}.
  Apply the Euler method to our equation
  of motion to solve for the analytic expression
  of position of a particle
  $y_n$ at the $nth$ timestep (i.e. Eq.~\ref{eq:euler}).

  In our simulations we use the timestep $\Delta t = 0.001$,
  examine Fig.~\ref{fig:0}
  for timesteps $\Delta t = 0.01$ and $\Delta t = 0.1$
  over the same maximum time
  (i.e. decrease the overall simulation time).
  How to the motions of the particle compare when the timestep
  becomes large?

@@

  Often molecular dynamics algorithms 
  are solved with higher order methods
  such as the Verlet method or Runge-Kutta methods.
  [{\it explain Verlet and/or RK4...}]
  Demonstrate how the Verlet method
  simplifies to the Euler method when
  the particle acceleration $\vec{a}=0$.

  {\it invite/guide students to play with the timestep and study the results in comparison to Euler method}

>>>>

  \item {\bf Integration Methods}
    \label{ex:euler}
    To calculate the position of the particle we
    integrate the equation of motion using
    the standard definition of velocity
    $\vec{v}_i = d\vec{r}_i/dt$ 
    via the 
    %In our model we use the
    Euler method. % to solve this equation.
    Our equation of motion provides
    a direct calculation for particle velocity.
    %
  When solving ordinary differential equations,
  the Euler method is effective for solving linear equations
  of the form
  $dy/dt = f(t,y(t))$ with initial condition $y(t_0) = y_0$.
  The solution is calculated algorithmically
  by stepping in time through $n$ integer steps
  $t_n = t_0 + n \Delta t$.
  At each subsequent step the new
  value for $y$ is calculated as a map solution using
  discrete times 
  $y_{n+1} = y_n + f(t_n,y_n)$.
  The Euler method can be applied to non-linear
  equations effectively if the time step $\Delta t$
  is kept sufficiently small \cite{Newman}.
  In our simulations we use the timestep $\Delta t = 0.001$,
  but we find for a single particle
  we can just as easily used 
  %examine 
  %for
  timesteps $\Delta t = 0.01$ and $\Delta t = 0.1$
  over the same maximum time
  (i.e. decrease the overall simulation time and sampling rate)
  and reproduce the motions shown in Fig.~\ref{fig:0}(b).
  This is due to the linear nature of the solution,
  and the lack of interaction with neighboring particles.
  Apply the Euler method to our equation
  of motion to solve for the analytic expression
  of position of a particle
  $y_n$ at the $nth$ timestep (i.e. Eq.~\ref{eq:euler}).
  %How to the motions of the particle compare when the timestep
  %becomes large?

@@

  Often molecular dynamics algorithms 
  are solved with higher order methods
  such as the Verlet method or Runge-Kutta methods.
  These methods include  higher order terms
  that provide accurate solutions for
  second order differential equations,
  i.e. $dy/dt = v$ and $dv/dt = f_2(y,t)$,
  where $f_2(y,t)$ is a second derivative function.
  In Newton's second law this would be proportional
  to the net force on a single particle.
  We do not integrate $f_2(y,t)$ since 
  $a = dv/dt = 0$,
  and thus rearrange $f_2(y,t)$ to solve directly for $v$,
  as described in Ex.~\ref{ex:n2l}.
  The Taylor expansion of the position function $y(t)$
  \begin{equation}
    y(t_n \pm \Delta t) = y(t_n) \pm \frac{dy}{dt} \Delta t + \frac{1}{2} \frac{d^2y}{dt^2} (\Delta t)^2 \pm O(\Delta t)^3 ... %\frac{1}{6} \frac{d^3y}{dt^3} (\Delta t)^3 + ...
  \end{equation}
  is exact at first order
  for the $n$-th time step.

  Demonstrate how the Verlet method
  simplifies to the Euler method when
  the particle acceleration $\vec{a}=0$.
